---
title: "Assignment 6"
output: html_document
date: "2026-03-02"
---

```{r setup, include=FALSE}
library(tidyverse)
library(rpart)
library(caret)
library(rpart.plot)
library(pROC)
library(kernlab)
library(NHANES)
library(e1071)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

set.seed(123) # EXTRA IMPORTANT 
```

1. Load the NHANES data using the NHANES R package

```{r}
data(NHANES) # initialize data

str(NHANES) # briefly examine what we have
```

2. Restrict the NHANES data to the list of 11 variables below. Partition the data into training and testing using a 70/30 split. Perform light data cleaning. Determine if you want to exclude any of the features before you start. 

"Age", "Race1", "Education", "HHIncome", "Weight", "Height", "Pulse", "Diabetes", "BMI", "PhysActive", "Smoke100"

```{r}
# filter data, clean names
# exclude weight and height due to multicollinearity with bmi
NHANES <- NHANES |> 
  janitor::clean_names() |>
  select(age, race1, education, hh_income, pulse, diabetes, bmi, phys_active, smoke100) |> drop_na() 

# Centering and Scaling
set.up.preprocess <- preProcess(NHANES , method = c("center", "scale"))

# Output pre-processed values
transformed.vals <- predict(set.up.preprocess, NHANES)

# Creating balanced partitions in the data
train.index <- createDataPartition(NHANES$diabetes, p = 0.7, list = FALSE)

nhanes.train <- NHANES[train.index, ]
nhanes.test <- NHANES[-train.index, ]

summary(NHANES$diabetes)
summary(nhanes.train$diabetes)
summary(nhanes.test$diabetes)
```

I excluded weight and height due to multicollinearity between the predictors with BMI. We split training and test data in 70/30 partitions by diabetes status. 

3. Construct three prediction models to predict diabetes using the features from NHANES. You will optimize each model using cross-validation to choose hyperparameters in the training data and then compare performance across models. You will use the following three algorithms to create your prediction models:

a) Classification Tree

```{r}
set.seed(123)

# 10-fold cross-validation and using down-sampling because of imbalance in data
train.control.class<-trainControl(method="cv", number=10, sampling="down")

# sequence of cp parameters to try 
grid.1<-expand.grid(cp=seq(0.001, 0.3, by=0.01))

# train model
tree.nhanes <-train(
                      diabetes~., 
                      data=nhanes.train, 
                      method="rpart",
                      trControl=train.control.class, 
                      tuneGrid=grid.1
                      )

tree.nhanes$bestTune
tree.nhanes

tree.nhanes$finalModel %>%
  rpart.plot()

# obtain variable importance on the final model within training data
varImp(tree.nhanes)

# accuracy metric and confusion matrix from training
confusionMatrix(tree.nhanes)
```

The CART model was trained with 10-fold cross-validation, yielding a final value of 0.001 for the complexity parameter $Cp$ and an average accuracy of 70.8% (based on the confusion matrix). 

b) Support Vector Classifier (i.e. Support Vector Machine with a linear classifier)

```{r}
set.seed(123)

# train model, note i already centered + scaled the data
# also set 10-fold cross-validation. for predicted probabilities, set class Probs=True
train_control<-trainControl(method="cv", number=10, sampling = "down", classProbs = T)

# incorporate different values for cost (C)
svm.caret <- train(
                    diabetes ~ ., 
                    data=nhanes.train, 
                    method="svmLinear",  
                    trControl=train_control, 
                    #preProcess=c("center", "scale"), 
                    tuneGrid=expand.grid(C=seq(0.001,2, length=30))
                    )

#Visualize accuracy versus values of C
plot(svm.caret)

#Obtain metrics of accuracy from training
confusionMatrix(svm.caret)

#See information about final model
svm.caret$finalModel
```

The Support Vector Machines with Linear Kernel was trained with 10-fold cross-validation, yielding a final value of 0.552 for the complexity parameter $Cp$ and an average accuracy of 72.4% (based on the confusion matrix). 

c) Logistic regression.

```{r}
set.seed(123)

train.control.lr <- trainControl(method="cv", number=10, sampling="down")

nhanes.lr <- train(
  diabetes ~ .,
  data = nhanes.train,
  method = "glm",
  family = "binomial",
  trControl = train.control.lr
)

nhanes.lr
```

The logistic regression model was trained with 10-fold cross-validation, yielding an average accuracy of 71.8%. 

4. Select an "optimal" model and calculate final evaluation metrics in the test set. ONLY ONE MODEL SHOULD BE APPLIED IN THE TEST SET What do you conclude about your final model's performance?

I'm going with the SVM w/ Linear Kernels because it was the most accurate (72.4%, compared to 70.8% for CART and 71.8% for LR) 

```{r}
set.seed(123)

# make predictions in testset
svm.pred.test <- svm.caret %>% 
              predict(nhanes.test)

#Get evaluation metrics from test set
confusionMatrix(svm.pred.test, nhanes.test$diabetes, positive="Yes")

#Create ROC Curve for Analysis
pred.prob <- svm.caret %>% 
  predict(nhanes.test, type = "prob")

#Create plot of Area under the Receiver Operating Curve (AUROC)
analysis <- roc(response=nhanes.test$diabetes, predictor=pred.prob[,2])
plot(1-analysis$specificities,analysis$sensitivities,type="l",
ylab="Sensitivity",xlab="1-Specificity",col="black",lwd=2,
main = "ROC Curve for Heart Disease Classification")
abline(a=0,b=1)

auc(analysis)
```

The final model, which uses SVM w/ Linear Kernels, has an accuracy of 73.4% and recall (Sensitivity) of 81.7% and True Negative Rate (Specificity) of 72.4%. With an AUC of 82.2, we can say this model does a good job differentiating between diabetic and non-diabetic individuals.

5. In this analysis, we've used Race as one of the predictors in the model. Briefly discuss the ethical considerations of including race in a disease prediction model. (no more than 1-2 paragraphs)

Race is potentially an unethical predictor based on the fact that it may be a poor approximation for the conditions that give rise to differential diabetes case prevalence across demographics. For example, race (Black, white, etc.) is probably less indicative of latent diabetic class assignment than say food access, zip-code, heritable, or pre-birth factors (such as gestational diabetes in the mother). While race may be a powerful predictor in the context of feature importance in our model, it does not reflect the entirety of, nor the best proxy for, the complex phenomena that we are interested in understanding.